An Alternative Geometric Interpretation of the Riemann-Stieltjes Integral

### 2017-12-15 paper.tex
finished Jonathans comments

\begin{abstract}
It is desirable to have geometric interpretations of difficult mathematical concepts. In this paper we present an intuitive geometric interpretation of the Riemann-Stieltjes integral. It can be 
summarized as follows: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small non-rectangular 
integration strips. This interpretation is surprisingly similar to the geometric interpretation of the ordinary Riemann integral. 
It is therefore likely that students, encountering the Riemann-Stieltjes integral for the first time, would find this new alternative geometric framework a useful visualization tool.
\end{abstract}

@@

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration
and in doing so developed the Cavalieri integral \cite{ackermann2012}. The area of a non-rectangular integration strip is computed via 
Cavalieri's principle \cite{andersen1985,eves1991, malik1984, wildberger2002, young1985}. They also showed that a Cavalieri integral can easily be converted to a Riemann-Stieltjes integral. 
However, the reverse operation was not described in \cite{ackermann2012}. In this paper we address this shortcoming. This 
reverse operation actually allows us to contrive a novel geometric interpretation of the Riemann-Stieltjes integral.
This geometric interpretation is: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small
non-rectangular integration strips. Interestingly enough, this interpretation is very similar 
to the geometric interpretation of the Riemann integral. The priciple difference between these interprations is the shape of the ingration strips; non-rectangular in the case of the Riemann-Stieltjies integral and rectangular in the case of the Rieman integral.  
This rectangular geometric interpretation has been quite useful when 
teaching the ordinary Riemann integral to undergraduate students \cite{jones2015,sealey2014}. Consquently, it seems likely that undergraduate students would find the non-rectangular geometric interpretation
of the Riemann-Stieltjes integral just as useful. It is also interesting to mention that there are a few papers in the literature that propose teaching simpler integral definitions (which are easier to grasp) before introducing students to the Riemann integral \cite{czarnocha2001,prabhu2008,ruffa2002,zaskis2014}. It therefore also seems plausible that
students will find it useful to first understand the simpler Cavalieri integral before they are exposed to the more abstract Riemann-Stieltjes integral.\\

@@

\begin{remark}
Note that the definition of the Cavalieri integral as presented above imposes much stricter requirements on $f$ and $g$ than the 
definition of the Riemann-Stieltjes integral does. To ensure consistency, we will assume, from here onwards and unless stated otherwise, that the stricter requirements imposed 
by the Cavalieri definition are now also binding on the Riemann-Stieltjes integral. Recall that the main aim of the paper is to 
help students develop a geometric intuition for the Riemann-Stieltjes integral that will be valid for most of the 
Riemann-Stieltjes integral problems that students will encounter during their undergraduate studies.
While this does not imply that this interpretation cannot provide insight into a broad range of problems, it does mean that the interpretation may not remain valid for each and every Riemann-Stieltjes integral
and not to present a complete geometric framework. This is not to say that the geometric interpretation presented here cannot be used to provide 
insight to a broad range of problems. In the example section we show that Cavalieri sums can provide meaningful insights even 
when we relax some of the strict restrictions imposed by the above definitions.
\end{remark}

@@

\noindent
\begin{remark}
It is interesting to note that G.~Bullock follows route similar to ours when developing his geometrical framework.
He starts of with quite strict requirements for $f$ ang $g$ and relaxes more and more of these requirements in his latter 
examples. 
\end{remark}

@@

\section{Alternative Geometric Interpretation}
\begin{figure}[htb]
\centering
\includegraphics[width=1\textwidth]{fig13_change}
\caption{A non-rectangular sum-based geometric interpretation of the Riemann-Stieltjes integral. Moreover, the figure depicts the region $R$ (also see Fig.~\ref{fig:ex1}). The integrator $g$ maps points 
in $[a',b']$ (depicted in blue) to points in $[a,b]$ (depicted in green). The function $h$ performs the reverse operation. Moreover, $g$ maps the $x$-coordinates at which
translations of $a(y)$ (depicted in red) intersect with $f(x)$ to the $x$-intercepts of the translated functions. This observation enables us to interpret the Riemann-Stieltjes integral as an infinite sum of infinitesimally small non-rectangular integration 
strips.}
\label{fig:2d_geo}
\end{figure}
%\begin{equation}
%\label{eq:cal_def_2D}
%\int_{a(y)}^{b(y)}fdx =\lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1,
%\end{equation}
%and that
%\begin{equation}
%\label{eq:rs_def_g}
%\int_{a'}^{b'} f \, dg(x) = \lim_{n\to \infty} \sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)].
%\end{equation}
%Fig.~\ref{fig:2d_geo} depicts a region whose area can be determined by considering non-rectangular integration strips.
%. Note that the shape of the integration strips are determined by $a(y)$. 
%In Fig.~\ref{fig:2d_geo}, the function $a(y)$ is depicted in red. If we translate $a(y)$ along the $x$-axis we actually create two partitions on the $x$-axis, 
%i.e. $(x_i^1)_{i=0}^{n}$ and $(x_i^2)_{i=0}^{n}$ with $x_i^1\in[a,b]$ and $x_i^2\in[a',b']$.  To map partition points in $[a',b']$ to partition points 
%in $[a,b]$ we use the function $g$. The reverse is achieved through $h$. The functions $g$ and $h$ depend soly on $a(y)$ and are also 
%computed with it.
Consider the definition of the Riemann-Stieltjes integral and 
Eq.~\eqref{eq:g_int}. For the sake of continiuty we repeat Eq.~\eqref{eq:g_int} below:
\begin{equation}
\label{eq:g_int2}
\int_{a'}^{b'} f(x) dg(x) =  \lim_{n \rightarrow \infty}\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)], 
\end{equation}
with $(x_i^2)_{i=0}^n$ being arbitrary partition points on the $x$-axis.
Eq.~\eqref{eq:g_int2} tells us that the function of the integrator $g$ is to map points in the interval $[a',b']$ to other 
locations on the $x$-axis. When, $g$ is monotone increasing it maps points in the interval $[a',b']$ to $[a,b]$,
with $a = g(a')$ and $b = g(b')$.\\

\noindent
To improve our understanding of $g$ lets investigate a specific example of the Riemann-Stieltjes integral in more depth: 
\begin{equation}
R = \int_{\frac{1}{2}}^1 x~d2x. 
\end{equation}
$R$ is depicted in Fig.~\ref{fig:2d_geo}. In this example $g$ maps $[\frac{1}{2},2]$ to $[1,4]$. The mapping between these two intervals is also 
depicted in Fig.~\ref{fig:2d_geo}. Moreover, the interval $[\frac{1}{2},2]$ is depicted in blue,
while the interval $[1,4]$ is depicted in green.\\

\noindent
Furthermore, consider $a(y) = 1-y$ (depicted in red in Fig.~\ref{fig:2d_geo}). Interestingly, the integrator $g$ maps the $x$-coordinates where translations of $a(y)$ and $f(x)$ intersect to the $x$-intercepts of the translations of $a(y)$.
The implications of this observation are the following: 
\begin{itemize}
 \item $g(x_{i+1}^2)-g(x_{i}^2)$ (i.e. $\Delta x_i^1$) represents the length of the base of a non-rectangular integration strip. Note, the base lengths in Fig.~\ref{fig:2d_geo} are all equal.
 \item $f(x_i^2)$ is the height of a non-rectangular integration strip.
 \item $f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$ is the area of a non-rectangular integration strip (due to Cavalieri's principle).
 \item The Riemann-Stieltjes sum, i.e. $\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$, is equal to the area obtained by summing together the area of multiple 
 integration strips.
 \item In the limit the Riemann-Stieltjes sum approaches $R$. 
\end{itemize}
The Riemann-Stieltjes integral can therefore be interpreted as the area which is obtained by summing together the areas of an infinite number of infinitesimally small non-rectangular integration strips,
i.e. a Riemann-Stieltjes integral can be converted into an equivalent Cavalieri integral.\\

@@

\begin{remark}
Note that the above procedure places an even stricter limitation on $f$, it now also needs to be invertible. Moreover the 
above derivation also requires that $a(0)$ exists.
\end{remark}

@@

\noindent
In Fig.~\ref{fig:ex1_sec} we plot the geometric interpretation of $\int_{1}^3 (2x+8)^3 d[x^2-20]$. This same figure also 
provides a geometric interpretation of $\int_{1}^3 (2x+8)^3 dx^2$ albeit a less intuitive one. The $a(y)$ function need not 
produce a mapping $g$ from $[1,3]$ to $[-19,-11]$. It can also produce a mapping $\widetilde{g}$ from $[1,3]$ to $[1,9]$. In both cases the 
depiction remains valid, however, in the case of $\widetilde{g}$ the lengths of the bases of the depicted integration strips are not 
computed on the $x$-axis itself, but rather on the line $y=1000$. The fact that the non-rectangular sum-based geometric framework remains useful even if $g(a')\neq a$ is true in general.\\

\noindent
In Fig.~\ref{fig:ex2_sec} we depict the geometric interpretation of $\int_1^2 \frac{1}{x} d[x^2+C]$. This example is a bit of a special case, since $a(0)$ does not exist. 
However, the non-rectangular sum-based geometric framework does still produce a meaningful result.\\

\noindent
One of the main limitations of the non-rectangular sum-based geometric interpretation we present in this paper is that both $f$ and $g$ need to be invertible.
Being invertible is however not that big of a limitation if we take the following into account. We can always split a Riemann-Stieltjes
integral whose integrand and integrator are not invertible over its integration limits into multiple constituent Riemann-Stieltjes integrals whose 
integrand and integrator are invertible over the integration limits of the constituent integrals. For example, say 
$f$ and $g$ are not invertible over the interval $[a',b']$, but are invertible over the intervals $[a',c']$ and 
$[c',b']$ then we may write:
\begin{equation}
\int_{a'}^{b'}f(x)dg(x) = \int_{a'}^{c'}f(x)dg(x) + \int_{c'}^{b'} f(x)dg(x).
\end{equation}
This allows us to to find a non-rectangular sum-based geometric interpretation of $\int_{a'}^{c'}f(x)dg(x)$ and $\int_{c'}^{b'} f(x)dg(x)$
which we can then depict in one image to give a single geometric interpretation of $\int_{a'}^{b'}f(x)dg(x)$.
In Fig.~\ref{fig:ex4_sec} we present just such a scenario. In this example $g(x)$ is not invertible over the interval 
$[0,2]$. It is however invertible over the interval $[0,1]$ and $[1,2]$. The geometric interpretation which we 
obtain by considering the intervals $[0,1]$ and $[1,2]$ separately is depicted, respectively, in the 
right and left panel of Fig.~\ref{fig:ex4_sec}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{EX2-crop.pdf}
\caption{The non-rectangular sum-based geometric interpretation of $\int_{\frac{-\pi}{2}}^0 \sin{x}~d[\cos{x}-1]$ (i.e this integral integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} \sin{x} dx$).}
\label{fig:ex2_sec}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{EX3-crop.pdf}
\caption{The non-rectangular sum-based geometric interpretation of $\int_1^2 \frac{1}{x} d[x^2+C]$ (i.e this integral integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} \frac{1}{x} dx$).}
\label{fig:ex3_sec}
\end{figure}

>>>>

\begin{abstract}
It is desirable to have geometric interpretations of difficult mathematical concepts. In this paper we present a new and more intuitive geometric interpretation of the Riemann-Stieltjes integral. It can be 
summarized as follows: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small non-rectangular 
integration strips. This interpretation is surprisingly similar to the geometric interpretation of the ordinary Riemann integral. 
It is therefore likely that students, when encountering the Riemann-Stieltjes integral for the first time, would find this new alternative geometric framework a useful visualization tool.
\end{abstract}

@@

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration
and in doing so developed the Cavalieri integral \cite{ackermann2012}. The area of a non-rectangular integration strip is computed via 
Cavalieri's principle \cite{andersen1985,eves1991, malik1984, wildberger2002, young1985}. They also showed that a Cavalieri integral can easily be converted to a Riemann-Stieltjes integral. 
However, the reverse operation was not described in \cite{ackermann2012}. In this paper we address this shortcoming. This 
reverse operation actually allows us to contrive a novel geometric interpretation of the Riemann-Stieltjes integral.
This geometric interpretation is: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small
non-rectangular integration strips. Interestingly enough, this interpretation is very similar 
to the geometric interpretation of the Riemann integral. The principle difference between these interprations is the shape of the ingration strips; non-rectangular in the case of the Riemann-Stieltjies integral and rectangular in the case of the Rieman integral.  
This rectangular geometric interpretation has been quite useful when 
teaching the ordinary Riemann integral to undergraduate students \cite{jones2015,sealey2014}. Consquently, it seems likely that undergraduate students would find the non-rectangular geometric interpretation
of the Riemann-Stieltjes integral just as useful. It is also interesting to mention that there are a few papers in the literature that propose teaching simpler integral definitions (which are easier to grasp) before introducing students to the Riemann integral \cite{czarnocha2001,prabhu2008,ruffa2002,zaskis2014}. It therefore also seems plausible that
students will find it useful to first understand the simpler Cavalieri integral before they are exposed to the more abstract Riemann-Stieltjes integral.\\

@@

\begin{remark}
Note that the definition of the Cavalieri integral as presented above imposes much stricter requirements on $f$ and $g$ than the 
definition of the Riemann-Stieltjes integral does. To ensure consistency, we will assume, from here onwards and unless stated otherwise, that the stricter requirements imposed 
by the Cavalieri definition are now also binding on the Riemann-Stieltjes integral. 
%Recall that the main aim of the paper is to 
%help students develop a geometric intuition for the Riemann-Stieltjes integral that will be valid for most of the 
%Riemann-Stieltjes integral problems that students will encounter during their undergraduate studies.
While this does not imply that this interpretation cannot provide insight into a broad range of problems, it does mean that the interpretation may not remain valid for each and every conceivable Riemann-Stieltjes integral.
In the example section we show that Cavalieri sums can provide meaningful insights even 
when we relax some of the strict restrictions imposed by the above definitions.
\end{remark}

@@

\noindent
\begin{remark}
It is interesting to note that G.~Bullock follows a route similar to ours when developing his geometrical framework.
He starts of with quite strict requirements for $f$ ang $g$ and relaxes more and more of these requirements in his latter 
examples. 
\end{remark}

@@

\section{Alternative Geometric Interpretation}
\begin{figure}[htb]
\centering
\includegraphics[width=1\textwidth]{fig13_change}
\caption{A non-rectangular sum-based geometric interpretation of the Riemann-Stieltjes integral. Moreover, the figure depicts the region $R$ (also see Fig.~\ref{fig:ex1}).
Its area is denoted by the symbol $A$. The integrator $g$ maps points 
in $[a',b']$ (depicted in blue) to points in $[a,b]$ (depicted in green). The function $h$ performs the reverse operation. Moreover, $g$ maps the $x$-coordinates at which
translations of $a(y)$ (depicted in red) intersect with $f(x)$ to the $x$-intercepts of the translated functions. This observation enables us to interpret the Riemann-Stieltjes integral as an infinite sum of infinitesimally small non-rectangular integration 
strips.}
\label{fig:2d_geo}
\end{figure}
%\begin{equation}
%\label{eq:cal_def_2D}
%\int_{a(y)}^{b(y)}fdx =\lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1,
%\end{equation}
%and that
%\begin{equation}
%\label{eq:rs_def_g}
%\int_{a'}^{b'} f \, dg(x) = \lim_{n\to \infty} \sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)].
%\end{equation}
%Fig.~\ref{fig:2d_geo} depicts a region whose area can be determined by considering non-rectangular integration strips.
%. Note that the shape of the integration strips are determined by $a(y)$. 
%In Fig.~\ref{fig:2d_geo}, the function $a(y)$ is depicted in red. If we translate $a(y)$ along the $x$-axis we actually create two partitions on the $x$-axis, 
%i.e. $(x_i^1)_{i=0}^{n}$ and $(x_i^2)_{i=0}^{n}$ with $x_i^1\in[a,b]$ and $x_i^2\in[a',b']$.  To map partition points in $[a',b']$ to partition points 
%in $[a,b]$ we use the function $g$. The reverse is achieved through $h$. The functions $g$ and $h$ depend soly on $a(y)$ and are also 
%computed with it.
Consider the definition of the Riemann-Stieltjes integral and 
Eq.~\eqref{eq:g_int}. For the sake of continuity we repeat Eq.~\eqref{eq:g_int} below:
\begin{equation}
\label{eq:g_int2}
\int_{a'}^{b'} f(x) dg(x) =  \lim_{n \rightarrow \infty}\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)], 
\end{equation}
with $(x_i^2)_{i=0}^n$ being arbitrary partition points on the $x$-axis.
Eq.~\eqref{eq:g_int2} tells us that the function of the integrator $g$ is to map points in the interval $[a',b']$ to other 
locations on the $x$-axis. When, $g$ is monotone increasing it maps points in the interval $[a',b']$ to $[a,b]$,
with $a = g(a')$ and $b = g(b')$.\\

\noindent
To further improve our understanding of $g$ lets investigate a specific example of the Riemann-Stieltjes integral in more depth: 
\begin{equation}
A =\int_{a'}^{b'} f(x) dg(x)=\int_{\frac{1}{2}}^1 x~d2x, 
\end{equation}
In this example $g$ maps $[\frac{1}{2},2]$ to $[1,4]$. The mapping between these two intervals is also 
depicted in Fig.~\ref{fig:2d_geo}. Moreover, the interval $[\frac{1}{2},2]$ is depicted in blue,
while the interval $[1,4]$ is depicted in green.\\

\noindent
Furthermore, consider $a(y) = 1-y$ (depicted in red in Fig.~\ref{fig:2d_geo}). Interestingly, the integrator $g$ maps the $x$-coordinates where translations of $a(y)$ and $f(x)$ intersect to the $x$-intercepts of the translations of $a(y)$.
The implications of this observation are the following: 
\begin{itemize}
 \item $g(x_{i+1}^2)-g(x_{i}^2)$ (i.e. $\Delta x_i^1$) represents the length of the base of a non-rectangular integration strip. Note, the base lengths in Fig.~\ref{fig:2d_geo} are all equal.
 \item $f(x_i^2)$ is the height of a non-rectangular integration strip.
 \item $f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$ is the area of a non-rectangular integration strip (due to Cavalieri's principle).
 \item The Riemann-Stieltjes sum, i.e. $\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$, is equal to the area obtained by summing together the area of multiple 
 integration strips.
 \item In the limit the Riemann-Stieltjes sum approaches $A$, i.e. the area of the region $R$ (depicted in Fig.~\ref{fig:2d_geo}). 
\end{itemize}
The Riemann-Stieltjes integral can therefore be interpreted as the area which is obtained by summing together the areas of an infinite number of infinitesimally small non-rectangular integration strips,
i.e. a Riemann-Stieltjes integral can be converted into an equivalent Cavalieri integral.\\

@@

\begin{remark}
Note that the above procedure places an even stricter limitation on $f$ as it now also needs to be invertible. Moreover, the 
above derivation also requires that $a(0)$ exists.
\end{remark}

@@

\noindent
In Fig.~\ref{fig:ex1_sec} we plot the geometric interpretation of $\int_{1}^3 (2x+8)^3 d[x^2-20]$. This same figure also 
provides a geometric interpretation of $\int_{1}^3 (2x+8)^3 dx^2$, albeit a less intuitive one. The $a(y)$ function need not 
produce a mapping $g$ from $[1,3]$ to $[-19,-11]$. It can also produce a mapping $\widetilde{g}$ from $[1,3]$ to $[1,9]$. In both cases the 
depiction remains valid. However, in the case of $\widetilde{g}$ the lengths of the bases of the depicted integration strips are not 
computed on the $x$-axis itself, but rather on the line $y=1000$. The fact that the non-rectangular sum-based geometric framework remains useful even if $g(a')\neq a$ is true in general.\\

\noindent
In Fig.~\ref{fig:ex2_sec} we depict the geometric interpretation of $\int_1^2 \frac{1}{x} d[x^2+C]$. This example is a bit of a special case, since $a(0)$ does not exist. 
However, the non-rectangular sum-based geometric framework still produces a meaningful result.\\

\noindent
One of the main limitations of the non-rectangular sum-based geometric interpretation we present in this paper is that both $f$ and $g$ need to be invertible.
Being invertible is, however, not a severe limitation if we take the following into account. We can always split a Riemann-Stieltjes
integral whose integrand and integrator are not invertible over its integration limits into multiple constituent Riemann-Stieltjes integrals whose 
integrand and integrator are invertible over their associated integration limits. For example, say 
$f$ and $g$ are not invertible over the interval $[a',b']$, but are invertible over the intervals $[a',c']$ and 
$[c',b']$ then we may write:
\begin{equation}
\int_{a'}^{b'}f(x)dg(x) = \int_{a'}^{c'}f(x)dg(x) + \int_{c'}^{b'} f(x)dg(x).
\end{equation}
This allows us to to find a non-rectangular sum-based geometric interpretation of $\int_{a'}^{c'}f(x)dg(x)$ and $\int_{c'}^{b'} f(x)dg(x)$
which we can then depict in one image to give a single geometric interpretation of $\int_{a'}^{b'}f(x)dg(x)$.
In Fig.~\ref{fig:ex4_sec} we present just such a scenario. In this example $g(x)$ is not invertible over the interval 
$[0,2]$. It is however invertible over the interval $[0,1]$ and $[1,2]$. The geometric interpretation which we 
obtain by considering the intervals $[0,1]$ and $[1,2]$ separately is depicted in the 
right and left panels of Fig.~\ref{fig:ex4_sec}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{EX2-crop.pdf}
\caption{The non-rectangular sum-based geometric interpretation of $\int_{\frac{-\pi}{2}}^0 \sin{x}~d[\cos{x}-1]$ (i.e this integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} \sin{x} dx$).}
\label{fig:ex2_sec}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{EX3-crop.pdf}
\caption{The non-rectangular sum-based geometric interpretation of $\int_1^2 \frac{1}{x} d[x^2+C]$ (i.e this integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} \frac{1}{x} dx$).}
\label{fig:ex3_sec}
\end{figure}
